---
output:
  pdf_document:
    includes:
      keep_tex: yes
    number_sections: yes
---

```{r settings, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, include=TRUE)
set.seed(1)
```
\newif\ifsol
\soltrue
\pagestyle{empty}
11\. The silicate Earth weighs approximately $4\times{10}^{24}$ kg and
contains 250 ppm potassium. How much ^40^Ar is created in 4.5 billion
years?  Given that ^40^Ar cannot escape from the Earth and the
quantity of ^40^Ar in the atmosphere is 66000 Gt, what is the ``K-Ar
age of the Earth's atmosphere''?  K has three isotopes with
^39^K/^40^K = 7771 and ^41^K/^40^K = 561. ^40^K is radioactive and
undergoes branched decay to ^40^Ar (10.72%) and ^40^Ca (89.28%) with a
half-life of 1.248 billion years. The molar mass of K is 39.1 g/mol.

   ```{r}
   GtEarth <- 4e24/1e12  # convert kg to Gt
   GtK <- GtEarth*250e-6 # Gt of K in the silicate Earth
   MK <- 39.1e-6 # Gt/Gmol
   GmolK <- GtK/MK # Gmol
   f40K <- 1/(7771 + 561 + 1) # 40K/K
   GmolK40 <- GmolK*f40K # Gmol of 40K in the silicate Earth
   GtAr40 <- 66e3
   M40K <- 40e-6 # Gt/Gmol
   GmolAr40 <- GtAr40/M40K
   Ar40K40 <- GmolAr40/GmolK40
   l40 <- log(2)/1.248
   fec <- 0.1072 # fraction of 40K decaying to 40Ar
   atmage <- log(Ar40K40/fec+1)/l40
   message("The 40Ar/40K-age of the Earth's atmosphere is ",
   signif(atmage,3),' Ga.')
   ```
