---
output:
  pdf_document:
    includes:
      keep_tex: yes
    number_sections: yes
---

```{r settings, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, include=TRUE)
set.seed(1)
```
\newif\ifsol
\soltrue
\pagestyle{empty}
14\. Analysis of a granite yields:
   ```{r include=FALSE}
   l38 <- IsoplotR::settings('lambda','U238')[1]
   l40 <- IsoplotR::settings('lambda','K40')[1]
   l87 <- IsoplotR::settings('lambda','Rb87')[1]
   tUPbzirc <- 520
   tKAramp <- 480
   tKArbio <- 400
   tRbSrmusc <- 470
   tRbSrbio <- 460
   PbUzirc <- exp(l38*tUPbzirc)-1
   ArKamp <- exp(l40*tKAramp)-1
   ArKbio <- exp(l40*tKArbio)-1
   SrRbmusc <- exp(l87*tRbSrmusc)-1
   SrRbbio <- exp(l87*tRbSrbio)-1
   ttab <- matrix(signif(c(PbUzirc,ArKamp,ArKbio,SrRbmusc,SrRbbio),3),1,5)
   colnames(ttab) <- c("^206^Pb*/^238^U (zircon)","^40^Ar*/^40^K (amphibole)",
   "^40^Ar*/^40^K (biotite)","^87^Sr*/^87^Rb (muscovite)",
   "^87^Sr*/^87^Rb (biotite)")
   ```
   ```{r include=TRUE, echo=FALSE, results='asis'}
   knitr::kable(ttab)
   ```
   What is the likely age of intrusion of this granite? How reliable
   is the result?

   ```{r}
   tt <- log(ttab+1)/c(l38,l40,l40,l87,l87)
   colnames(tt) <- c('U-Pb (zirc)','K-Ar (amp)','K-Ar (bio)',
   'Rb-Sr (musc)','Rb-Sr (bio)')
   print(tt)
   ```
   \ifsol
   The true age is probably around 550--520 Ma. The reason
   being that some disruptive phenomenon has affected the region as
   there is clearly a discordance in ages. The most robust of the
   chronometers is U-Pb on zircons, but it is not perfect and this
   apparent age may be made a little older. But this reasoning is
   very rough and ready! A more detailed assessment would involve
   textural analysis, diffusion experiments and numerical modelling.
   \fi
