---
output:
  pdf_document:
    includes:
      keep_tex: yes
    number_sections: yes
---

```{r settings, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, include=TRUE)
set.seed(1)
```
\newif\ifsol
\soltrue
\pagestyle{empty}
16\. Consider a rapidly cooled (volcanic) apatite of 10 Ma with an
atomic Th/U ratio of 5. What is the ratio of helium atoms to fission
tracks in it?  Look up the relevant decay constants in the notes.

   \ifsol
   The helium ingrowth equation:

   $$
   He =
   \left[
   8 \frac{137.818}{138.818} (e^{\lambda_{238}t}-1) +
   7 \frac{1}{138.818} (e^{\lambda_{235}t}-1)
   \right] U + 6 (e^{\lambda_{232}t}-1) Th
   $$

   Because Th/U=5:

   $$
   He =
   \left[
   8 \frac{137.818}{138.818} (e^{\lambda_{238}t}-1) +
   7 \frac{1}{138.818} (e^{\lambda_{235}t}-1) +
   5 \times 6 (e^{\lambda_{232}t}-1)
   \right] U
   $$

   The fundamental fission track age equation:

   $$
   n_s = \frac{\lambda_f}{\lambda_{238}}
   \frac{137.818}{138.818} U (e^{\lambda_{238}t}-1)
   $$

   Hence:

   $$
   \frac{He}{n_s} =
   \frac{
      8 \frac{137.818}{138.818} (e^{\lambda_{238}t}-1) +
      7 \frac{1}{138.818} (e^{\lambda_{235}t}-1) +
      30 (e^{\lambda_{232}t}-1)
   }{
      \frac{\lambda_f}{\lambda_{238}}
      \frac{137.818}{138.818} (e^{\lambda_{238}t}-1)
   }
   $$

   In \texttt{R}:
   \fi
   ```{r}
   lf <- 8.46e-11 # Myr-1
   l238 <- 1.55125e-4 # Myr-1
   l235 <- 9.8485e-4 # Myr-1
   l232 <- 4.95e-5 # Myr-1
   tap <- 10 # Ma
   num <- 8*(137.818/138.818)*(exp(l238*tap)-1) +
   (7/138.818)*exp(l235*tap-1) + 30*(exp(l232*tap)-1)
   den <- (lf/l238)*(137.818/138.818)*(exp(l238*tap)-1)
   HeFT <- num/den
   message('The He/FT-ratio is ',signif(HeFT,2))
   ```

   \ifsol
   Even though spontaneous fission of ${}^{238}$U happens `only'
   1.8 million times less frequently than $\alpha$-decay,
   the apatite has produced 55 million times more $\alpha$ particles
   than fission tracks because (1) each ${}^{238}$U atom produces
   not one but eight helium atoms, and (2) not only ${}^{238}$U
   but also ${}^{235}$U and ${}^{232}$Th contribute to the helium
   budget. NB: $\alpha$-recoil damages the crystal lattice of
   apatite and other U-bearing minerals, thereby affecting its
   diffusivity. Thus the closure temperature of apatite can
   vary as a function of time and U (and Th) concentration.
   \fi
